---
title: "Assignment 3: Data Exploration"
author: "Andrew Brantley, Section #1"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}

#checking working directory
getwd()

#loading necessary packages
library("tidyverse")

#uploading datasets
Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)

Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Understanding to what degree different types of insects are affected could be important in knowing if target pest insects are being eliminated or if non-target insects are being affected. For example, if non-target pollinators are being affected by the insecticide this would be counterproductive to farmers working to produce more crop so ensuring that target species are the only ones being affected would be vital information for users of the insecticide.

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Especially in Western states such as Colorado, forest litter that falls to the ground will often eventually become fuel for forest fires. Tracking the amount of litter being deposited in different areas of forest would be important information in modeling/predicting fire location, size, and intensity in future fire seasons as well as tracking current fires. 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
* Falling litter and woody debris are monitored through both elevated PVC traps and ground traps
* The sampling takes place in areas with woody vegetation >2 meters tall
* Trap placements can be either targeted or randomized depending on the vegetation within the plot



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}

#dimensions of Neonics dataset
dim(Neonics)

#other summaries
head(Neonics)
tail(Neonics)
str(Neonics)
length(Neonics)

```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}

#summary of Effect column
summary(Neonics$Effect)

```

> Answer: The most commonly studied effects are mortality (1493) and population (1803). These are likely the most studied effects because they are the target result of these insecticides as they are meant to kill off insects that may be harming crops, lowering the poopulation.

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}

#summary of common names
summary(Neonics$Species.Common.Name)

```

> Answer: The six most commonly studied insects are as follows: Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and Italian Honey Bee. These insects are largely responsible for pollination of plants around the globe making them of high interest as they are liekly not the target of these insecticide applications. 

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}

#determining class of Conc.1..Author.
class(Neonics$Conc.1..Author.)

```

> Answer: The class of Conc.1..Author. is factor and it is not numeric becuase of the argument we used when loading in the dataset "stringsAsFactors = TRUE". This column also features a variety of units which are paired with them in the next column over "Conc.Units.1..Author.".



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}

#plot of number of studies each year (by publication)
ggplot(Neonics, aes(x=Neonics$Publication.Year)) +
  geom_freqpoly()

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}

#plot of number of studies each year (by publication) with color based on Test.Location
ggplot(Neonics, aes(x=Publication.Year, color = Test.Location)) +
  geom_freqpoly()

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The large majority of publications are now lab-based with field natural being the next most common. Prior to 2010 most publications in most years were actually field natural based with lab-based being the next most common.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}

#barplot of end point counts
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar()

```

> Answer: LOEL and NOEL are mcuh more common than any of the other endpoints. LOEL is the endpoint representing lowest-observable-effect-level and NOEL represents the endpoint of no-observable-effect-level.


## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}

#checking class of collectDate
class(Litter$collectDate)

#changing collectDate class from factor to date
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")

# confirming new class, confirmed to be Date
class(Litter$collectDate)

#which dates litter was sampled in August 2018, determined to be "2018-08-02" "2018-08-30"
unique(Litter$collectDate)

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

#how many plots sampled at Niwot Ridge
unique(Litter$plotID)


```

> Answer: There are 12 different plots that were sampled at Niwot Ridge. Information obtained with the `unique` function returns each of the unique values in that column/vector while `summary` provides the all the unique values but also how many times each of them appear in the column/vector.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}

#bar graph of functionalGroup counts
ggplot(Litter, aes(functionalGroup)) +
  geom_bar()

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}

#box plot of dryMass
ggplot(Litter, aes(x=Litter$functionalGroup, y=Litter$dryMass)) +
  geom_boxplot()

#violin plot of dryMass
ggplot(Litter, aes(x=Litter$functionalGroup, y=Litter$dryMass)) +
  geom_violin()

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The box plot more clearly shows the mean dryMass for each litter type as well as outliers that exist. On the other hand, the violin plots poorly represent spread within the data as they end up appearing as lines due to some dramatic outliers and the majority of the data being very low dryMass. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: The litter type with the highest biomass at these sites is needles with the next highest litter types being mixed and twigs/branches. All other litter types have fairly negligible dry masses. 

