---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Andrew Brantley"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1

# setting up workspace, loading data
getwd()

library(tidyverse)
library(agricolae)
library(ggplot2)
library(lubridate)

Lake.Data <- read.csv(file = "../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",
                      stringsAsFactors = TRUE)

#setting date column as date
Lake.Data$sampledate <- as.Date(Lake.Data$sampledate, format = "%m/%d/%y", sep = "/")

#2

# building personal theme
Andrew.Theme <- theme_gray(base_size = 14) +
  theme(axis.text = element_text(colour = "black", face = "italic"), 
        legend.position = "right", 
        panel.grid.major.x = element_line(colour = "black", linetype = 3, size = 0.5), 
        panel.grid.major.y =  element_line(colour = "black", linetype = 3, size = 0.5))

theme_set(Andrew.Theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Lake temperature in July does not change with depth across all lakes.
Ha: Lake temperature in July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

# wrangling dataset
Lake.Data.subset <- Lake.Data %>% 
  mutate(month = month(sampledate)) %>% # adding month column
  filter(month == 7) %>% # filtering for July collections
  select(lakename:daynum, depth, temperature_C) %>% # selecting columns
  filter(!is.na(temperature_C)) # eliminating NAs

#5

# visualizing temperature and depth relationship
DepthTempPlot <- ggplot(Lake.Data.subset, aes(depth, temperature_C)) +
                    geom_point() +
                    geom_smooth(method = lm, color = "blue") +
                    labs(title = "Temperature by Depth", x = "Depth", 
                         y = "Temperature (C)") +
                    ylim(0,35) +
                    Andrew.Theme
DepthTempPlot


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This plot suggests that temperature decreases as depth increases. The distribution of points is much more variable closer to the surface and asymptotes as the depth increases along with being less variable.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7 

# performing linear regression

Depth.Temp.Regression <- lm(data = Lake.Data.subset, temperature_C ~ depth)
summary(Depth.Temp.Regression)

# displaying results

par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(Depth.Temp.Regression)


```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: According to the linear regression 73.9% of variability in temperature is explained by depth on 9726 degrees of freedom. The statistical significance of this result is well below the 0.05 level commonly used as it comes out to be <2.2e-16. Temperature is expected to drop by 1.95 degrees Celsius per 1 m drop in depth. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

# running AIC
Depth.Temp.AIC <- lm(data = Lake.Data.subset, 
                     temperature_C ~ year4 + daynum + depth)
step(Depth.Temp.AIC)

#10

# running multiple regression
TempModel <- lm(data = Lake.Data.subset, 
                temperature_C ~ year4 + daynum + depth)
summary(TempModel)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: Removing any of the three predictor variables increases the AIC so all three are included in the multiple regression (year4, daynum, depth). This model explains 74.1% of the variability in temperature. This is a slight improvement from 73.9% from the linear model with jsut depth as a predictor variable.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

# ANOVA with aov function
Temp.ANOVA <- aov(data = Lake.Data.subset, temperature_C ~ lakename)
summary(Temp.ANOVA)

# ANOVA with lm function
Temp.ANOVA2 <- lm(data = Lake.Data.subset, temperature_C ~ lakename)
summary(Temp.ANOVA2)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: With both methods of completing this analysis there are significant differences in mean temperatures among all of the lakes.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

DepthTemp.Lake.Plot <- ggplot(data = Lake.Data.subset, 
                              aes(depth, temperature_C, 
                                  color = lakename)) + # including color here 
                                          #so each lake gets a regression line
                          geom_point(alpha = 0.5) +
                          geom_smooth(method = "lm", se = FALSE) +
                          labs(title = "Temeperature by Depth For Each Lake", 
                               x = "Depth", y = "Temperature (C)") +
                          scale_colour_discrete("Lake Name") +
                          ylim(0,35) +
                          Andrew.Theme
DepthTemp.Lake.Plot


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

# Tukey's HSD test

Temp.Tukey <- TukeyHSD(Temp.ANOVA)
Temp.Tukey

Temp.Tukey.groups <- HSD.test(Temp.ANOVA, "lakename", group = TRUE)
Temp.Tukey.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: According to the grouping analysis of the ANOVA Ward Lake and Paul Lake have means that are statistically the same as Peter Lake. Every lake has a mean that is statistically the same as at least one other lake, none of them are statistically different from all the others.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A two sample t-test could be used to determine if there is a significant difference between the mean temperatures of Peter Lake and Paul Lake. The null hypothesis of such a test would be that there is no difference in the means while the alternative hypothesis would be that they are statistically different.


