---
title: "Assignment 09: Data Scraping"
author: "Andrew Brantley"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1

# checking working directory
getwd()

# loading packages
library(tidyverse)
library(rvest)
library(lubridate)
library(zoo)

# building personal theme
Andrew.Theme <- theme_gray() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(Andrew.Theme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
# creating webpage object
WaterSupply_Website <- 
  read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <-   WaterSupply_Website %>% 
                        html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
pswid <-               WaterSupply_Website %>% 
                        html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
ownership <-           WaterSupply_Website %>% 
                        html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
max.withdrawals.mgd <- WaterSupply_Website %>% 
                        html_nodes('th~ td+ td') %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4

# creating dataframe from vectors
Durham2020_Water <- data.frame(
              "Water System Name" = water.system.name,
              "PSWID" = pswid,
              "Ownership" = ownership,
              "Max Withdrawals (mgd)" = max.withdrawals.mgd,
              "Month" = as.factor(c(01, 05, 09, 02, 06, 10, 03, 07, 11, 04, 08, 12)))
Durham2020_Water

#5

# plotting max withdrawals 
MaxWithdrawals.Plot <- ggplot(Durham2020_Water, aes(x = Month, y = Max.Withdrawals..mgd.)) +
                          geom_line(group = 1) +
                          labs(title = "Durham 2020 Max Daily Withdrawals (mgd)")
MaxWithdrawals.Plot


```

6. Note that the PSWID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PSWID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.

# creating function to scrape website
WaterUsage_Scraper <- function(the_pswid, the_year){
 
  #Get the proper url
  the_url <- paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
                      the_pswid, '&year=',the_year)
  
#Scrape the data
water.system.name <-   read_html(the_url) %>% 
                        html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
pswid <-               read_html(the_url) %>% 
                        html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
ownership <-           read_html(the_url) %>% 
                        html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
max.withdrawals.mgd <- read_html(the_url) %>% 
                        html_nodes('th~ td+ td') %>% html_text()
  
#Convert to dataframe
WaterUsage_DF <- data.frame(
              "Water System Name" = as.character(water.system.name),
              "PSWID" = as.character(pswid),
              "Ownership" = as.character(ownership),
              "Max Withdrawals (mgd)" = as.character(max.withdrawals.mgd),
              "Month" = as.factor(c(01, 05, 09, 02, 06, 10, 03, 07, 11, 04, 08, 12)),
              "Year" = the_year)
  
#Return the dataframe
return(WaterUsage_DF)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7

# pulling 2015 max withdrawals for Durham
Durham2015 <- WaterUsage_Scraper("03-32-010", 2015)
Durham2015

# plotting Durham 2015 max withdrawals
Durham2015.Plot <- ggplot(data = Durham2015) +
                          geom_line(aes(Durham2015$Month, 
                                        Durham2015$Max.Withdrawals..mgd.), group = 1)
Durham2015.Plot


```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8

Asheville2015 <- WaterUsage_Scraper("01-11-010", 2015)
Asheville2015

# combining datasets
avl_dur_2015 <- bind_rows(Durham2015, Asheville2015)
avl_dur_2015

avl.dur.plot <- ggplot(avl_dur_2015) +
                    geom_point(aes(x = Month, y = Max.Withdrawals..mgd., color = PSWID)) +
                    labs( title = "Durham and Asheville 2015 Max Withdrawal Comparison")
avl.dur.plot

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9

# gathering and plotting Asheville withdrawal 2010-2019
years <- seq(2010,2019)
Avl <-   "01-11-010"
Avl_2010_2019 <- map2(Avl, years, WaterUsage_Scraper) %>%  bind_rows() %>% 
                      mutate("MonthYear" = make_date(year = Year, month = Month), 
                             "Max.Withdrawals..mgd." = as.numeric(Max.Withdrawals..mgd.))
Avl_2010_2019

Avl_2010_2019.Plot <-  ggplot(Avl_2010_2019, aes(x = MonthYear, y = Max.Withdrawals..mgd.)) +
                          geom_point() +
                          geom_line() +
                          geom_smooth(method='loess', se = FALSE) +
                          labs(title = "Asheville Max Withdrawals (mgd) 2010-2019", 
                               y = "Monthly Max Withdrawals (mgd)", x = "Date")
Avl_2010_2019.Plot

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
It seems that there has been an increase in water usage over time, especially in the last two years. 
